Owl occupancy in three of El Salvador's protected areas from 2003 to 2013

### 2021-01-12 manuscript_working/ms_working_RMD.Rmd
introduction edits

Owls, as a keystone species, are an important, yet cryptic, bioindicator of ecosystem health. Top predators also often have long-lasting and wide-ranging effects through trophic cascades. The various spotted owl sub-species (*Strix occidentalis*) of North America have long been studied as an indicator species for the interactions between anthropogenic and environmental forces, demonstrating that even various sub-species of spotted owl have complex and contrasting behavior responses to environmental land use and management [e.g., @White:2013; @Wan:2018].

Neotropical owls are less well-studied than temperate species, such as the spotted owls of North America, and we have limited understanding of their ecological requirements, population dynamics, reproductive behavior, or conservation status [@Enriquez:2006; @Clark:1978; @Perez-Leon:2017; @Rangel-Salazar:2017]. Yet, neotropical owls are reliant upon ecosystems under tremendous anthropogenic pressure and rapid and unpredictable responses to climate change [@Corlett:2012]. In the tropics, particularly in rain and cloud forests, owl studies are more difficult than in temperate regions, although there are more species living in the same habitat [@Konig:1999]. In general, while our knowledge about the conservation status of neotropical populations is limited, we do know that owl abundance and distribution is decreasing in several regions, where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

The vocalization of crepuscular and cryptic animals such as owls are often leveraged for surveying, and vocal patterns are often considered more important than coloration of plumage as an aid to identification [@Springer:1978; @Forsman:1983; @Konig:1999]. *springer and forsman not cited on doc from Jane xxx* Owls vocalize to communicate with the same species and to delimit territory [@Johnsgard:1988].  Spontaneous and provocation listening are two methods for auditory surveys that have been used with increasing frequency in surveys of nocturnal birds of prey [e.g., @Baumgardt:2019].  Spontaneous or passive surveys are based off of listening for animal calls with no prior stimulus whereas provocation surveys are conducted by listening after a period of playing animal vocalizations or "broadcast calls." One benefit of auditory surveys (including both spontaneous and broadcast) is being able to identify multiple species simultaneously. Adding in provocation to auditory surveys can increase the detection probability of owls, although the intra- and interspecific responses to broadcast calls could play a role in response patterns [@Enriquez:1997;@Baumgardt:2019].

The objective of our study was to monitor population dynamics of El Salvador owls through time to determine both individual species' population dynamics and species community composition. El Salvador is located in the heart of the Mesoamerican Biodiversity Hotspot [@Myers:2000].   We surveyed owls on foot using passive listening and broadcast vocalizations in two transects each within the three largest protected areas of El Salvador from 2003 through 2013.

This project was carried out during the breeding season (March-May) for three weeks each year over a 10-year period in the three previously mentioned protected natural areas.  These areas were selected to represent different ecosystems found in the country and the transect routes were selected to represent the habitats in the protected natural areas.

>>>>

Owls are a keystone species in many ecosystems, and as such are an important bioindicator of ecosystem health. Top predators such as owls affect communities through their top-down control on biodiversity and their presence or absence can cause trophic cascades. However, owls are also cryptic, nocturnal predators that are difficult to accurately census. The Mexican Spotted Owl (*Strix occidentalis*) and its sub-species have long been studied as an indicator species for the interactions between anthropogenic and environmental forces, demonstrating that even various sub-species of spotted owls have complex and often contrasting responses to environmental land use and management [e.g., @White:2013; @Wan:2018].

Neotropical owls are less well-studied than temperate species, and we have limited understanding of their ecological requirements, population dynamics, reproductive behavior, or conservation status [@Enriquez:2006; @Clark:1978; @Perez-Leon:2017; @Rangel-Salazar:2017]. Yet, neotropical owls are reliant upon ecosystems under tremendous anthropogenic pressure and rapid and unpredictable responses to climate change [@Corlett:2012]. In the tropics, particularly in rain and cloud forests, owl studies are even more difficult than in temperate regions, regardless of denser populations [@Konig:1999]. In general, while our knowledge about the conservation status of neotropical populations is limited, we do know that owl abundance and distribution is decreasing in several regions where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

The vocalization of crepuscular and cryptic animals such as owls are often leveraged for surveying, and vocal patterns are often considered more important than coloration as an aid to identification [@Springer:1978; @Forsman:1983; @Konig:1999]. *springer and forsman not cited on doc from Jane xxx* Owls vocalize to communicate with the same species and to delimit territory [@Johnsgard:1988].  Spontaneous and provocation listening are two methods that researchers use with increasing frequency to survey nocturnal birds of prey [e.g., @Baumgardt:2019].  Spontaneous or passive surveys are based off of listening for animal calls with no prior stimulus whereas provocation surveys are conducted by listening after a period of playing animal vocalizations or "broadcast calls." One benefit of auditory surveys (including both spontaneous and broadcast) is being able to identify multiple species simultaneously. Adding in provocation to auditory surveys can increase the detection probability of owls, although intra- and inter-species interactions can play a role whether owls are more likely or less likely to call in response to a vocalization simulus [@Enriquez:1997;@Baumgardt:2019].

The objective of our study was to monitor population dynamics of El Salvador owls through time to determine both individual species' population dynamics and community composition. El Salvador is located in the heart of the Mesoamerican Biodiversity Hotspot [@Myers:2000], and we selected the three survey areas to represent ecosystem diversity across El Salvador. We set up two routes in each protected area to further represent the habitats diversity of this heterogeneous country. We conducted nighttime foot surveys using passive listening and broadcast calls during the breeding season for three weeks each year from from 2003 through 2013.
